---
title: "Understanding demographic variations in health insurance coverage in the United States: evidence from ADD Health"
subtitle: "EDLD 651 Final Paper"
author: "Sara Kiros, Ohnmar Khine Min, Chen Su"
bibliography: references.bib
link-citations: true
format: 
  html:
    toc: true
    code-overflow: scroll
    css: styles.css
execute: 
  eval: true
  echo: false
  warning: false
fig-width: 6.5
fig-height: 8
---



```{r}
#| label: load-packages-raw-data

library(tidyverse)
library(here)
library(rio)
library(janitor)
library(gt)
library(dplyr)
library(ggplot2)
library(scales)
library(forcats)


pwave5 <- import(here("data", "pwave5.xpt")) |>
  characterize() |>
  clean_names() 


```



```{r}
#| label: select-variables

wave5_raw <- pwave5 |>
     select(aid, imonth5, iyear5, h5od1m, h5od1y,
            h5od2a, h5od4a, h5od4b, h5od4c,
            h5od4d, h5od4e, h5od4f, h5od4g, h5od8,
            h5lm5, h5lm6, h5lm7, h5lm9, h5ec2,
            h5id9, h5id10a, h5id10b, h5id10c, h5id10d)

```



```{r}
#| label: clean-employment

#Job Type
job_labels <- c(
  "1" = "Military", 
  "2" = "Private for-profit", 
  "3" = "Private nonprofit",
  "4" = "Local government",
  "5" = "State government", 
  "6" = "Federal government", 
  "7" = "Self-employed (non-inc)",
  "8" = "Self-employed (inc)",
  "9" = "Unpaid family worker"
  )


wave5_clean <- wave5_raw |> 
    mutate(
    employed = ifelse(h5lm5 == 1, "Employed",
                      ifelse(h5lm5 %in% c(2, 3), "Not employed", NA)),
    num_jobs = ifelse(h5lm6 %in% 1:4, h5lm6, NA),
    hours_work = ifelse(h5lm7 >= 5 & h5lm7 <= 90, h5lm7, NA),
     job_type = job_labels[as.character(h5lm9)]
)

wave5_clean$job_type <- as.factor(wave5_clean$job_type)
wave5_clean$employed <- as.factor(wave5_clean$employed)

```



```{r}
#| label: clean-sex

wave5_clean <- wave5_clean |> 
  mutate(
    sex = ifelse(h5od2a == 1, "Male", 
                    ifelse(h5od2a == 2, "Female", NA))
  )

#sex_employment
sex_employment <- wave5_clean |> 
  group_by (sex, employed) |>
  summarize(n = n(), .groups = "drop")

# removing missing

sex_employment <- na.omit(sex_employment)

```



```{r}
#| label: clean-insurance-status

wave5_clean <- wave5_clean %>%
 mutate(insurance_status = case_when(h5id9 == 1 ~ "Insurance Through Work",
                                     h5id9 == 2 ~ "Insurance Through Union",
                                     h5id9 == 4 ~ "Partner's Insurance",
                                     h5id9 == 6 ~ "Active Duty Military Insurance",
                                     h5id9 == 7 ~ "Self Private Insurance",
                                     h5id9 == 8 ~ "Self Marketplace or ACA Insurance",
                                     h5id9 == 9 ~ "Medicaid",
                                     h5id9 == 10 ~ "Medicare",
                                     h5id9 == 11 ~ "Department of Veterans Affairs",
                                     h5id9 == 12 ~ "Military Healthcare Plans",
                                     h5id9 == 13 ~ "Indian Health Service",
                                     h5id9 == 14 ~ "Unknown Health Insurance",
                                     h5id9 == 15 ~ "No Health Insurance"
                                     ))
  
wave5_clean$insurance_status <- as.factor(wave5_clean$insurance_status)

# no insurance reason 
wave5_clean <- wave5_clean %>%
 mutate(notinsured_reason = case_when(h5id10a == 1 ~ "Not offered via work or school",
                                     h5id10b == 1 ~ "Too expensive",
                                     h5id10c == 1 ~ "Do not need or want insurance",
                                     h5id10d == 1 ~ "Denied insurance"))


wave5_clean$notinsured_reason <- as.factor(wave5_clean$notinsured_reason)

```



```{r}
#| label: clean-income

#Household Income in the last calendar year (h5ec2)

wave5_clean <- wave5_clean %>%
  mutate(
    household_income_raw = factor(
      h5ec2,
      levels = 1:13,
      labels = c(
        "Less than $5,000",
        "$5,000 to $9,999",
        "$10,000 to $14,999",
        "$15,000 to $19,999",
        "$20,000 to $24,999",
        "$25,000 to $29,999",
        "$30,000 to $39,999",
        "$40,000 to $49,999",
        "$50,000 to $74,999",
        "$75,000 to $99,999",
        "$100,000 to $149,999",
        "$150,000 to $199,999",
        "$200,000 or more"
      )
    )
  )


# collapsing income 

wave5_clean<- wave5_clean %>%
  mutate(
    household_income = fct_collapse(household_income_raw,
       `Below 50k` = c("Less than $5,000", "$5,000 to $9,999", "$10,000 to $14,999", "$15,000 to $19,999","$20,000 to $24,999",
      "$25,000 to $29,999",
      "$30,000 to $39,999",
      "$40,000 to $49,999"),
      `50k-100k` = c("$50,000 to $74,999",
        "$75,000 to $99,999"),
      `100k-200k` = c("$100,000 to $149,999",
        "$150,000 to $199,999"),
      `Above 200k` = c("$200,000 or more")))


```



```{r}
#| label: convert-date-variables

survey_date_string <- paste(wave5_clean$iyear5, wave5_clean$imonth5, 1, sep = "-")

wave5_clean <- wave5_clean |>
  mutate(survey_date = ymd(survey_date_string))


birth_date_string <- paste(wave5_clean$h5od1y, wave5_clean$h5od1m, 1, sep = "-")

wave5_clean <- wave5_clean |>
  mutate(birth_date = ymd(birth_date_string))

```



```{r}
#| label: calculate-age-at-survey

wave5_clean <- wave5_clean |>
  mutate(age = interval(birth_date, survey_date) |>
           time_length(unit = "years") |> round(digits = 2))



```



```{r}
#| label: convert-race-ethnicity-variables

wave5_clean <- wave5_clean |>
  mutate(race = case_when(h5od4a == 1 & h5od4b == 0 & h5od4d == 0 & h5od4e == 0 &
                            h5od4f == 0 & h5od4g == 0 ~ "White", 
                          h5od4a == 0 & h5od4b == 1 & h5od4d == 0 & h5od4e == 0 &
                            h5od4f == 0 & h5od4g == 0 ~ "Black", 
                          h5od4a == 0 & h5od4b == 0 & h5od4d == 1 & h5od4e == 0 &
                            h5od4f == 0 & h5od4g == 0 ~ "Asian", 
                          h5od4a == 0 & h5od4b == 0 & h5od4d == 0 & h5od4e == 1 &
                            h5od4f == 0 & h5od4g == 0 ~ "Pacific Islander", 
                          h5od4a == 0 & h5od4b == 0 & h5od4d == 0 & h5od4e == 0 &
                            h5od4f == 1 & h5od4g == 0 ~ "American Indian or Alaska Native",
                          h5od4a == 0 & h5od4b == 0 & h5od4d == 0 & h5od4e == 0 &
                            h5od4f == 0 & h5od4g == 1 ~ "Other Race"))

# NA = multiple race selected, recode to multiracial
wave5_clean <- wave5_clean |>
  mutate(race = replace_na(race, "Multiracial"))


# ethnicity
wave5_clean <- wave5_clean |>
  mutate(ethnicity = case_when(h5od4c == 1 ~ "Hispanic",
                               h5od4c == 0 ~ "Non-Hispanic"))

# NA = Missing for ethnicity
wave5_clean <- wave5_clean |>
  mutate(ethnicity = replace_na(ethnicity, "Missing"))
    
# re-level variables
wave5_clean$race <- factor(wave5_clean$race,levels = c("White",
                                                       "Black",
                                                       "Asian",
                                                       "Pacific Islander",
                                                       "American Indian or Alaska Native",
                                                       "Multiracial",
                                                       "Other Race"))

wave5_clean$ethnicity <- factor(wave5_clean$ethnicity,
                                levels = c("Non-Hispanic",
                                           "Hispanic",
                                           "Missing"))

```



```{r}
#| label: race-ethnicity-recoding

# using pivot_wider to combine race/ethnicity
wave5_clean <- wave5_clean |> 
  pivot_wider(names_from = race,
              values_from = ethnicity)

# recoding into mono vs multi racial and ethnicity
wave5_clean <- wave5_clean |>
  mutate(`Non-Hispanic Multiracial` = case_when(Multiracial == "Non-Hispanic"~ 1),
         `Hispanic Multiracial` = case_when(Multiracial == "Hispanic"~ 1),
         `Non-Hispanic Monoracial` = case_when(is.na(Multiracial) & h5od4c == 0  ~ 1),
         `Hispanic Monoracial` = case_when(is.na(Multiracial) & h5od4c == 1  ~ 1))

# recoding NA to 0 (no)
wave5_clean <- wave5_clean |>
  mutate(`Non-Hispanic Multiracial` = replace_na(`Non-Hispanic Multiracial`, 0),
         `Hispanic Multiracial` = replace_na(`Hispanic Multiracial`, 0),
         `Non-Hispanic Monoracial` = replace_na(`Non-Hispanic Monoracial`, 0),
         `Hispanic Monoracial` = replace_na(`Hispanic Monoracial`, 0))

# pivot_longer to one variable on race/ethnicity combined  
wave5_clean <- wave5_clean |>
  pivot_longer(cols = `Non-Hispanic Multiracial`:`Hispanic Monoracial`,
               names_to = "race_ethnicity",
               values_to = "selected"
               ) 

# remove extra variables and rows
wave5_clean <- wave5_clean |> 
  filter(selected == 1) |> 
  select(-selected, -Multiracial, -White, -Black, -Asian, -`Pacific Islander`, 
         -`American Indian or Alaska Native`, -`Other Race`)

# factor level
wave5_clean$race_ethnicity <- factor(wave5_clean$race_ethnicity,
                                levels = c("Non-Hispanic Monoracial",
                                           "Non-Hispanic Multiracial",
                                           "Hispanic Monoracial",
                                           "Hispanic Multiracial"))

```




**to dos**
Add in methods section how you cleaned the data for each variable you cleaned.
Add ADD Health citation (make sure it renders and shows) throughout the write-up where ADD Health is referenced.
Add in the summary or key takeaways of finding (doesn't have to include everything in our results/discussion) in abstract.
Double check citation lists and making sure everything renders properly (make sure we are using "sex" and not gender since the question ask for biological sex assigned at birth, no code showing, all graphs displaying, in-line code working and citation working in-text and at the end)
Check structure of paper (does the table and graph referenced come right after the write-up description? are the table and figures numbered correctly?). Proofread for typos, sentences that don't make sense; check we hit all the requirement of rubric.
Delete all of our notes for communicating with each other in this doc before submitting (including this to-do).

NOTE: MAKE SURE THE NUMBER OF CODE CHUNK LABELS ARE NOT REPEATS, otherwise it will not render.
ALSO - model after the format of the first table to make sure we have subtitles of data source, and not using count but rather n of participants (n, or participants(n) or number of participants depending on the context/aesthetics)

@sara - could you make one table or graph on income -> health insurance? and also write up the description of the finding and reference the table/figure.

Here's the factor variable for income now:
household_income = 
       `Below 50k` ,
      `50k-100k`,
      `100k-200k`,
      `Above 200k`.
      
@ohnmar - I don't think we need both a table and a graph for job type and health insurance status (i think we should stick with your graph cuz it's cool). For all the other tables/figures without health insurance status as a variable, i think you can add an inline code and description just saying we found that majority of participants work for private for-profit etc. see an example of the in-line code I have for age and number of participants), same with employment and sex, can describe it in written sentences with in-line code.



## Abstract
Health insurance coverage is an important factor for individuals residing in the United States to access a variety of preventative and treatment services in medical settings [@HealthyPeople2030]. The current study examines the variations of health insurance status across demographic characteristics including race, ethnicity, age, sex, employment, income using Wave V of ADD Health survey data. [summarize finding here: We found that....delete me once this is done] --> We found that lower-income classes had higher rated of being uninsured and the primary provider of health insurance was throught the workplace across all income levels. These findings highlights the importance of addressing disparities in health insurance coverage to improve access to care for diverse population, which is crucial for the successful prevention and management of chronic health conditions for all.


## Introduction

Access to health services has been identified as a social determinant of health by the Office of Disease Prevention and Health Promotion [@HealthyPeople2030]. Status on health insurance coverage is a factor that largely contributes to accessing healthcare services, for both treatment and preventative services. Many adverse outcomes can result from a lack of health insurance coverage. Uninsured adults are less likely to receive preventative screenings for chronic conditions, such as cancer, cardiovascular disease, and diabetes, resulting in higher mortality among uninsured groups due to untimely diagnoses and treatments [@IOM2009]. In addition, there’s a higher likelihood of medical debt among uninsured groups due to accrued out-of-pocket expenses [@Herman2011]. 

Access to health insurance coverage is unequally distributed across people with lower incomes, racial/ethnic minorities,gender minorities, and young adults in the United States [@Hadley2003]. Particularly, disparities across health insurance coverage primarily affect people in lower-income classes or minoritized racial and ethnic groups, as about half of the uninsured population in the U.S. are also in lower-income classes or identified as belonging to a minoritized racial and ethnic group [@Majerol2015]. As the U.S. population has become more racially and ethnically diverse and more individuals identified as multiracial [@Jones2021], it is crucial to understand whether there are any variations in monoracial and multiracial individuals' access to health insurance and services. The current study aims to examine the distribution of health insurance coverage and how this may differ by race (monoracial, multiracial), ethnicity, age, sex, income and employment using ADD Health Wave 5 survey data.


## Methods

The current study uses the public available version of the ADD Health Wave 5 survey data, which was collected from 2016 through 2018. As the full dataset contains variables on many social, environmental, behavior and biological measures beyond the scope of the current study, we only selected variables that are relevant to our aim as a first step of our data cleaning process. 
We used the following R packages [@R; @here;@tidyverse;@rio;@janitor;@gt] to assist our data cleaning, as well as [@ggplot2;@scales;@forcats;@dplyr] for table and visualization creations of our results. We used [@addhealth] to download the raw data.

As we are interested in examining the potential differences in health insurance coverage by comparing monoracial and multiracial individuals that either identifies as Hispanic or Non-Hispanic rather than across each racial and ethnic group, we decided to exclude participants who have missing data on race and ethnicity. Therefore, our final sample includes `r nrow(wave5_clean)` adult participants (Mean age =`r round(mean(wave5_clean$age), 2)` years, Range = `r round(min(wave5_clean$age), 0) ` - `r round(max(wave5_clean$age), 0)` years).


Additionally, we are interested in examining differences in status of health insurance and types of health insurance across income levels. The original raw data had thirteen income levels, starting at `"Less than $5,000"`, increasing in increments of `$5,000` up to `"$200,000 or more"`. Therefore, we decided to collapse some of these levels to aid clarity of visualizations with less levels. We reorganized the original income levels to`Below 50k`,`50k-100k`,`100k-200k`, and`Above 200k`.

```{r}
#| label: race-health-table

#descriptive table

desc_table_demo <- wave5_clean |> 
  filter(!is.na(sex), !is.na(employed),
         !is.na(household_income)) |> 
  group_by (sex, race_ethnicity, employed, household_income) |>
  summarize(n = n(), .groups = "drop")

desc_table_demo |>
    mutate(percent = n / sum(n)) |>
    arrange(desc(n)) |>
    gt() |> 
    tab_header(title = "Table 1: Descriptive Statistics of Sex, Racial Ethnic Group, Employment, Household Income",
               subtitle = "Data Source: Wave 5 of ADD Health") |> 
    cols_label(
      sex = "Sex", 
      race_ethnicity = "Racial Ethnic Group", 
      employed = "Employment Status",
      household_income = "Household Income",
      n = "Participants (n)", 
      percent = "Percentage"
    ) |> 
    fmt_percent(columns = percent, decimals = 1)


```



```{r}
#| label: paper-figure-race

race_ethnicity <- wave5_clean |> 
  filter(!is.na(insurance_status)) |>
  group_by (race_ethnicity, insurance_status) |>
  summarize(n = n(), .groups = "drop") |>
  mutate(insurance_status = reorder(insurance_status, n))

race_ethnicity |>
  ggplot(aes(x = insurance_status, y = n, fill = race_ethnicity)) +
  geom_col(position = "dodge") + 
  labs(
    title = "Figure 1: Distribution of Health Insurance Coverage by Racial Ethnic Group", 
    subtitle = "Data Source: Wave 5 of ADD Health",
    caption = "Data Source: Wave 5 of ADD Health",
    x = "Insurance Coverage Type", 
    y = "Number of Participants", 
    fill = "Racial Ethnic Group"
  ) +
  theme_minimal() + coord_flip()

```



```{r}
#| label: paper-table-1

# descriptive table of gender and employment
sex_employment_table <- sex_employment|> 
  group_by(sex) |> 
  mutate(
    percent = n / sum(n) 
  ) |> 
  ungroup() |> 
  rename(count = n)
  

sex_employment_table |> 
  gt() |> 
  tab_header(title = "Table 2: Employment Status by Sex",
             subtitle = "Data Source: Wave 5 of ADD Health") |> 
  cols_label(
    sex = "Sex", 
    employed = "Employment Status",
    count = "Count", 
    percent = "Percent"
  ) |> 
  fmt_percent(columns = percent, decimals = 1)

```



```{r}
#| label: paper-figure-1
#Employment Status by Gender
ggplot(sex_employment_table, aes(x = sex, y = count, fill = employed)) +
  geom_col(position = "dodge") + 
  labs(
    title = "Figure 2: Employment Status by Sex",
    subtitle = "Data Source: Wave 5 of ADD Health",
    x = "Sex", 
    y = "Count", 
    fill = "Employment Status"
  ) + 
  theme_minimal()
```



```{r}
#| label: paper-figure-2
#Insurance and Employment visualization

insurance_by_employment <- wave5_clean |> 
  filter(!is.na(employed), !is.na(insurance_status)) |> 
  group_by(employed, insurance_status) |> 
  summarise(count = n(), .groups = "drop")

ggplot(insurance_by_employment, aes(
  y = fct_reorder(insurance_status, count), 
  x = count, 
  fill = employed)) +
  geom_col(position = "dodge") +
  labs(
    title = "Figure 3: Health Insurance Coverage by Employment Status",
    subtitle = "Data Source: Wave 5 of ADD Health",
    x = "Count", 
    y = "Insurance Type", 
    fill = "Employment"
  ) + 
  theme_minimal()
```


```{r}
#| label: paper-table-2

# descriptive table of job type.

jobtype_table <- wave5_clean |> 
  filter(!is.na(job_type)) |> 
  count(job_type, name = "count") |> 
  mutate(percent = count/ sum(count))

jobtype_table |> 
  gt() |> 
  tab_header(title = "Table 3: Distribution of Job Types Among Participants", 
             subtitle = "Data Source: Wave 5 of ADD Health") |> 
  cols_label(
    job_type = "Job Type", 
    count = "Count", 
    percent = "Percent"
  ) |> 
  fmt_percent(columns = percent, decimals = 1)
```



```{r}
#| label: paper-figure-3

#Health Insurance Status by Job Type
jobtype_insurance <- wave5_clean |> 
  filter(!is.na(job_type), !is.na(insurance_status)) |> 
  count(job_type, insurance_status) |> 
  group_by(job_type) |> 
  mutate(percent = n / sum(n)) |> 
  ungroup()

#Health Insurance Status by Job Type

jobtype_insurance <- wave5_clean |> 
  filter(!is.na(job_type), !is.na(insurance_status)) |> 
  count(job_type, insurance_status) |> 
  group_by(job_type) |> 
  mutate(percent = n / sum(n)) |> 
  ungroup()

jobtype_insurance |> 
  gt() |> 
  tab_header(title = "Table 4:Health Insurance Status by Job Type", 
             subtitle = "Data Source: Wave 5 of ADD Health") |> 
  cols_label(
    job_type = "Job Type", 
    insurance_status = "Health Insurance", 
    n = "Count", 
    percent = "Percent"
  ) |> 
  fmt_percent(columns = percent, decimals = 1)

```


```{r}
#| label: paper-figure-4
#Figure: Health Insurance Coverage by Job Type 


jobtype_insurance |>  ggplot(aes(
  x = fct_reorder(job_type, percent, .fun = max), 
  y = percent, 
  fill = insurance_status
)) + 
  geom_col() +
  coord_flip() +
  labs(
    title = "Figure 4: Health Insurance Distribution by Job Type", 
    subtitle = "Data Source: Wave 5 of ADD Health",
    x = "Job Type", 
    y = "Percent Within Job Type", 
    fill = "Insurance Type"
  ) + 
  theme_minimal() + 
 scale_y_continuous(
   breaks = seq(0, 1, 0.2),
   labels = c("0%", "20%", "40%", "60%", "80%", "100%")
 )


```


```{r}
#| label: paper-figure-5
#Figure: Proportions of Insurance Types for each Income Level

wave5_clean <- wave5_clean %>%
  mutate(
    household_income = fct_relevel(household_income,
                                   "Below 50k", "50k-100k", "100k-200k", "Above 200k"),
    insurance_status = fct_relevel(insurance_status,
      "Insurance Through Work",
      "Insurance Through Union",
      "Partner's Insurance",
      "Active Duty Military Insurance",
      "Self Private Insurance",
      "Self Marketplace or ACA Insurance",
      "Medicaid",
      "Medicare",
      "Department of Veterans Affairs",
      "Military Healthcare Plans",
      "Indian Health Service",
      "Unknown Health Insurance",
      "No Health Insurance"
    )
  )

pal <- c(
  "Insurance Through Work" = "#1b9e77",
  "Insurance Through Union" = "#66a61e",
  "Partner's Insurance" = "#7570b3",
  "Active Duty Military Insurance" = "#e6ab02",
  "Self Private Insurance" = "#a6761d",
  "Self Marketplace or ACA Insurance" = "#666666",
  "Medicaid" = "#e7298a",
  "Medicare" = "#d95f02",
  "Department of Veterans Affairs" = "#3288bd",
  "Military Healthcare Plans" = "#5e3c99",
  "Indian Health Service" = "#4daf4a",
  "Unknown Health Insurance" = "#b2abd2",
  "No Health Insurance" = "#d73027"
)

stack_df <- wave5_clean %>%
  filter(!is.na(household_income), !is.na(insurance_status)) %>%
  count(household_income, insurance_status, name = "n") %>%
  group_by(household_income) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

ggplot(stack_df,
       aes(x = household_income, y = pct, fill = insurance_status)) +
  geom_col(width = 0.8) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = pal, name = "Insurance Status") +
  labs(
    title = "Figure 5: Insurance Status Distribution by Income Level",
    subtitle = "Data Source: Wave 5 of ADD Health",
    x = "Household Income Levels",
    y = "Types of Insurance & Corresponding Proportions"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 15),
    axis.title = element_text()
  ) 


#citation("dplyr")
#@ggplot2,@scales,@forcats,@dplyr
# Make citation for library(dplyr)
# library(ggplot2)
# library(scales)
# library(forcats)
# and Add Health Data 
#@addhealth
```




```{r}
#| label: paper-table-
# descriptive table of No Health Insurance and reasons why.

noinsure_table <- wave5_clean |> 
  filter(!is.na(notinsured_reason)) |> 
  count(notinsured_reason, name = "count") |> 
  mutate(percent = count/ sum(count))

noinsure_table |> 
  gt() |> 
  tab_header(title = "Table 5: Distribution of Reasons behind No Health Insurance",
             subtitle = "Data Source: Wave 5 of ADD Health") |> 
  cols_label(
    notinsured_reason = "Job Type", 
    count = "Count", 
    percent = "Percent"
  ) |> 
  fmt_percent(columns = percent, decimals = 1)


```



## Results

We presented descriptive statistics of our demographic variables of interest in Table 1 below. The majority of our participants are employed, reported their household income as above 50k and their sex assigned at birth as female, and identified as Non-Hispanic Monoracial (see Table 1).The primary reason behind not having health insurance was it being too expensive at 55.1% (see Table 5). The largest proportion of insurance types across the four income levels was receiving health insurance through the workplace (see Figure 5). However, the proportions of receiving insurance through the workplace was lower at both the lowest and highest end of income levels, `Below 50k` and `Above 200k`. The second leading insurance type was "Medicaid" for those in the `Below 50k` category, and "Partner's Insurance" in the rest of the categories. 

[sara ohnmar please summarize the finding of your table/visualizations here and have the interpretation in discussion]





## Discussion

### Findings about Gender and Employment
The descriptive table 1 and figures 1 highlight differences in employment status by gender and how it is linked to health insurance coverage. In this sample and analysis of Gender and Employment table, women have higher raw counts of employment because the female sample is larger.However, men shows the higher employment rate (87.3%) compared to women (79.6%). Women also have a higher count of "not employed" but proportionally still lower than men. Since there are small NA counts, it suggests that there will be minimal missing data and do not affect the patterns that much. This table of Gender and Employment status reflects labor-force disparities in U.S. It suggests slight gender differences in employment stability or participation rates.

The figure 1: Employment Status by Gender visualizes the same pattern that both genders are mostly employed. The bar chart visually shows larger absolute employment counts for women. However, proportionally, men are more likely to be employed when sample size is considered. It indicates gender differences in labor market participation in this sample. This analysis highlights the importance of distingushing between absolute counts and percentage. Employment is tightly linked to employer-sponsored insurance, so even small gender differences can influence coverage patterns. 

### Findings about Income levels 
Figure 5 revealed that receiving health insurance through the workplace remains the primary reason across all income levels. In addition, the largest proportion of "No Health Insurance" responses was in the lowest income level, however was partially mitigated by Medicaid. These figure findings highlight the importance in altering employer reliant healthcare insurance models, as federally provided health insurance, such as Medicaid, can clearly aid in drastically reducing uninsured statuses. 


### Findings about Reasons for No Health Insurance 
Table 5 revealed that the primary reason behind not having health insurance was due to it being too expensive or costly (55.1%). These results shed light on how cost/affordability continues to be major barrier to health insurance status and accessing health services. The second leading reason of not having health insurance, "Not offered via work or school"(38.5%), convey the risks from relying heavily on workplaces to provide health insurance, which is the gap of leaving unemployed persons at a higher risk of being uninsured, based on their employment status. These results can be used to support universal healthcare policies by highlighting existing health care disparities stemming from these flawed healthcare systems.

### Implications for Health Insurance
Higher male employment rates explain why men may have greater access to employer-based health insurance. Since women have relatively lower employment rate, it may increase reliance on non-employer insurance options such as Medicaid and partner insurance. This is because U.S. insurance is heavily employment-dependent, gendered labor-market patterns can translate directly into coverage disparities. These findings align with broader literature on women's higher likelihood of part-time work or employment interruptions which may limit access to employer benefits.

### Findings about Job Type
Job types vary significantly across participants and private for-profit employment is the most common one. Government jobs such as local, state and federal also represent meaningful shares of the sample. Self-employed and unpaid family workers represents smaller but important groups.

### Insurance Access Patterns
Private for-profit and government job types show the highest rates of employer-sponsored insurance while self-employed individuals have much lower access to employer insurance and rely more on individual market plans or remin uninsured. Unpaid or informal job types show very low employer-insurance coverage and higher Medicaid and uninsured rates. Therefore, job type is a strong structural determinant of insurance access even among people who are technically "employed". 

### Implications
Employer-provided insurance is unequally distributed across job sectors. Workers in traditional and stable employment sectors such as governments or for-profit benefit from higher coverage. Workers in self-employment or informal roles face systemic disadvantages in getting decent insurance. Unemployed persons face higher risks of having no health insurance coverage due to systemic reliance on employer-sponsored health insurance models. Those with lower incomes also face higher risks for having no health insurance coverage, thus, leading to higher risks of delaying health services or prioritizing preventative care. This support the argument that the structure of the labor market creates disparities in health insurance access beyond employment status alone. 

# References



