---
title: "Understanding demographic variations in health insurance coverage in the United States: evidence from wave 5 of ADD Health survey"
subtitle: "EDLD 651 Final Paper"
author: "Sara Kiros, Ohnmar Khine Min, Chen Su"
bibliography: references.bib
link-citations: true
format: 
  html:
    toc: true
execute: 
  eval: true
  echo: false
  warning: false
fig-width: 6.5
fig-height: 8
---



```{r}
#| label: load-packages-raw-data

#install.packages("webshot2")
#install.packages("labeled")
library(tidyverse)
library(here)
library(rio)
library(janitor)
library(gt)
library(labelled)
library(webshot2)

pwave5 <- import(here("data", "pwave5.xpt")) |>
  characterize() |>
  clean_names() 

```



```{r}
#| label: select-variables

wave5_raw <- pwave5 |>
     select(aid, imonth5, iyear5, h5od1m, h5od1y,
            h5od2a, h5od4a, h5od4b, h5od4c,
            h5od4d, h5od4e, h5od4f, h5od4g, h5od8,
            h5lm5, h5lm6, h5lm7, h5lm9, h5ec2,
            h5id9, h5id10a, h5id10b, h5id10c, h5id10d)

```



```{r}
#| label: clean-employment

#Job Type
job_labels <- c(
  "1" = "Military", 
  "2" = "Private for-profit", 
  "3" = "Private nonprofit",
  "4" = "Local government",
  "5" = "State government", 
  "6" = "Federal government", 
  "7" = "Self-employed (non-inc)",
  "8" = "Self-employed (inc)",
  "9" = "Unpaid family worker"
  )


wave5_clean <- wave5_raw |> 
    mutate(
    employed = ifelse(h5lm5 == 1, "Employed",
                      ifelse(h5lm5 %in% c(2, 3), "Not employed", NA)),
    num_jobs = ifelse(h5lm6 %in% 1:4, h5lm6, NA),
    hours_work = ifelse(h5lm7 >= 5 & h5lm7 <= 90, h5lm7, NA),
     job_type = job_labels[as.character(h5lm9)]
)

wave5_clean$job_type <- as.factor(wave5_clean$job_type)
wave5_clean$employed <- as.factor(wave5_clean$employed)

```



```{r}
#| label: clean-sex

wave5_clean <- wave5_clean |> 
  mutate(
    sex = ifelse(h5od2a == 1, "Male", 
                    ifelse(h5od2a == 2, "Female", NA))
  )

#sex_employment
sex_employment <- wave5_clean |> 
  group_by (sex, employed) |>
  summarize(n = n(), .groups = "drop")

# removing missing

sex_employment <- na.omit(sex_employment)

```



```{r}
#| label: clean-insurance-status

wave5_clean <- wave5_clean %>%
 mutate(insurance_status = case_when(h5id9 == 1 ~ "Insurance Through Work",
                                     h5id9 == 2 ~ "Insurance Through Union",
                                     h5id9 == 4 ~ "Partner's Insurance",
                                     h5id9 == 6 ~ "Active Duty Military Insurance",
                                     h5id9 == 7 ~ "Self Private Insurance",
                                     h5id9 == 8 ~ "Self Marketplace or ACA Insurance",
                                     h5id9 == 9 ~ "Medicaid",
                                     h5id9 == 10 ~ "Medicare",
                                     h5id9 == 11 ~ "Department of Veterans Affairs",
                                     h5id9 == 12 ~ "Military Healthcare Plans",
                                     h5id9 == 13 ~ "Indian Health Service",
                                     h5id9 == 14 ~ "Unknown Health Insurance",
                                     h5id9 == 15 ~ "No Health Insurance"
                                     ))
  
wave5_clean$insurance_status <- as.factor(wave5_clean$insurance_status)

# no insurance reason 
wave5_clean <- wave5_clean %>%
 mutate(notinsured_reason = case_when(h5id10a == 1 ~ "Not offered via work or school",
                                     h5id10b == 1 ~ "Too expensive",
                                     h5id10c == 1 ~ "Do not need or want insurance",
                                     h5id10d == 1 ~ "Denied insurance"))


wave5_clean$notinsured_reason <- as.factor(wave5_clean$notinsured_reason)

```



```{r}
#| label: clean-income

#Household Income in the last calendar year (h5ec2)

wave5_clean <- wave5_clean %>%
  mutate(
    household_income_raw = factor(
      h5ec2,
      levels = 1:13,
      labels = c(
        "Less than $5,000",
        "$5,000 to $9,999",
        "$10,000 to $14,999",
        "$15,000 to $19,999",
        "$20,000 to $24,999",
        "$25,000 to $29,999",
        "$30,000 to $39,999",
        "$40,000 to $49,999",
        "$50,000 to $74,999",
        "$75,000 to $99,999",
        "$100,000 to $149,999",
        "$150,000 to $199,999",
        "$200,000 or more"
      )
    )
  )


# collapsing income 

wave5_clean<- wave5_clean %>%
  mutate(
    household_income = fct_collapse(household_income_raw,
       `Below 50k` = c("Less than $5,000", "$5,000 to $9,999", "$10,000 to $14,999", "$15,000 to $19,999","$20,000 to $24,999",
      "$25,000 to $29,999",
      "$30,000 to $39,999",
      "$40,000 to $49,999"),
      `50k-100k` = c("$50,000 to $74,999",
        "$75,000 to $99,999"),
      `100k-200k` = c("$100,000 to $149,999",
        "$150,000 to $199,999"),
      `Above 200k` = c("$200,000 or more")))


```



```{r}
#| label: convert-date-variables

survey_date_string <- paste(wave5_clean$iyear5, wave5_clean$imonth5, 1, sep = "-")

wave5_clean <- wave5_clean |>
  mutate(survey_date = ymd(survey_date_string))


birth_date_string <- paste(wave5_clean$h5od1y, wave5_clean$h5od1m, 1, sep = "-")

wave5_clean <- wave5_clean |>
  mutate(birth_date = ymd(birth_date_string))

```



```{r}
#| label: calculate-age-at-survey

wave5_clean <- wave5_clean |>
  mutate(age = interval(birth_date, survey_date) |>
           time_length(unit = "years") |> round(digits = 2))

wave5_clean |>
  ggplot(aes(x=age)) + geom_histogram(binwidth = 1,
                                      fill="skyblue", color="white", alpha=0.8) + 
  labs(x = "Age at Wave 5",
       y = "Number of Participants",
    title = "Distribution of Participant's Age",
       caption = "Data Source: Wave 5 ADD Health") + 
  theme_minimal()


```



```{r}
#| label: convert-race-ethnicity-variables

wave5_clean <- wave5_clean |>
  mutate(race = case_when(h5od4a == 1 & h5od4b == 0 & h5od4d == 0 & h5od4e == 0 &
                            h5od4f == 0 & h5od4g == 0 ~ "White", 
                          h5od4a == 0 & h5od4b == 1 & h5od4d == 0 & h5od4e == 0 &
                            h5od4f == 0 & h5od4g == 0 ~ "Black", 
                          h5od4a == 0 & h5od4b == 0 & h5od4d == 1 & h5od4e == 0 &
                            h5od4f == 0 & h5od4g == 0 ~ "Asian", 
                          h5od4a == 0 & h5od4b == 0 & h5od4d == 0 & h5od4e == 1 &
                            h5od4f == 0 & h5od4g == 0 ~ "Pacific Islander", 
                          h5od4a == 0 & h5od4b == 0 & h5od4d == 0 & h5od4e == 0 &
                            h5od4f == 1 & h5od4g == 0 ~ "American Indian or Alaska Native",
                          h5od4a == 0 & h5od4b == 0 & h5od4d == 0 & h5od4e == 0 &
                            h5od4f == 0 & h5od4g == 1 ~ "Other Race"))

# NA = multiple race selected, recode to multiracial
wave5_clean <- wave5_clean |>
  mutate(race = replace_na(race, "Multiracial"))


# ethnicity
wave5_clean <- wave5_clean |>
  mutate(ethnicity = case_when(h5od4c == 1 ~ "Hispanic",
                               h5od4c == 0 ~ "Non-Hispanic"))

# NA = Missing for ethnicity
wave5_clean <- wave5_clean |>
  mutate(ethnicity = replace_na(ethnicity, "Missing"))
    
# re-level variables
wave5_clean$race <- factor(wave5_clean$race,levels = c("White",
                                                       "Black",
                                                       "Asian",
                                                       "Pacific Islander",
                                                       "American Indian or Alaska Native",
                                                       "Multiracial",
                                                       "Other Race"))

wave5_clean$ethnicity <- factor(wave5_clean$ethnicity,
                                levels = c("Non-Hispanic",
                                           "Hispanic",
                                           "Missing"))

```


**remove at the end before submitting**
```{r}
#| label: visual-race-ethnicity

wave5_clean |>
  ggplot(aes(race)) + geom_bar(fill = "darkgreen") + 
  labs(title = "Distribution of Participant's Race and Ethnicity",
       caption = "Data Source: Wave 5 ADD Health") + 
  theme_minimal() + facet_wrap(~ethnicity) + coord_flip()

# just race
wave5_clean |>
  ggplot(aes(race)) + geom_bar(fill = "darkgreen") + 
  labs(title = "Distribution of Participant's Race",
       caption = "Data Source: Wave 5 ADD Health") + 
  theme_minimal() + coord_flip()

# just ethnicity
wave5_clean |>
  ggplot(aes(ethnicity)) + geom_bar(fill = "darkorange") + 
  labs(title = "Distribution of Participant's Ethnicity",
       caption = "Data Source: Wave 5 ADD Health") + 
  theme_minimal() + coord_flip()

# table
 wave5_clean |>
  count(race, ethnicity) |>
  group_by(ethnicity) |>
  gt() |>
  tab_header(title = "Descriptive Statistics: Race and Ethnicity") |>
  cols_label(
    race = "Race",
    ethnicity = "Ethnicity"
    )


```




## Abstract
[1-2 Sentences about study rationale --> Access to health services has been identified as a social determinant of health by the Office of Disease Prevention and Health Promotion [@HealthyPeople2030]. Status on health insurance coverage is a factor that largely contributes to accessing healthcare services, for both treatment and preventative services. 


The current study examines the variations of health insurance status across demographic characteristics including race, ethnicity, age, sex, employment, income using Wave V of ADD Health survey data. [summarize finding here: We found that....]. It is critical to address disparities in health insurance status as improved access to care for diverse population is crucial for the successful prevention and management of chronic health conditions for all.

## Introduction
(1-2 paragraphs) Describe here why we are examining these demographic variables and health insurance outcome.

Many adverse outcomes can result from a lack of health insurance coverage. Uninsured adults are less likely to receive preventative screenings for chronic conditions, such as cancer, cardiovascular disease, and diabetes, resulting in higher mortality among uninsured groups due to untimely diagnoses and treatments [@IOM2009]. In addition, there’s a higher likelihood of medical debt among uninsured groups due to accrued out-of-pocket expenses [@Herman2011]. Access to health insurance coverage is unequally distributed across people with lower incomes, racial/ethnic minorities,gender minorities, and young adults [@Hadley2003]. Particularly, disparities across health insurance coverage primarily affect people in lower-income classes or in minority groups, as about half of the uninsured population in the U.S. are also in lower-income classes or in a minority group [@Majerol2015].


## Methods

The current study uses the public available version of the ADD Health Wave 5 survey data, which was collected from 2016 through 2018. The sample includes `r nrow(wave5_clean)` adult participants.

```{r}
#| label: paper-table-1

# descriptive table of gender and employment
gender_employment_table <- sex_employment|> 
  group_by(sex) |> 
  mutate(
    percent = n / sum(n) 
  ) |> 
  ungroup() |> 
  rename(count = n)
  

gender_employment_table |> 
  gt() |> 
  tab_header(title = "Employment Status by Gender") |> 
  cols_label(
    sex = "Gender", 
    employed = "Employment Status",
    count = "Count", 
    percent = "Percent"
  ) |> 
  fmt_percent(columns = percent, decimals = 1)

```

```{r}
#| label: paper-figure-1
#Employment Status by Gender
ggplot(gender_employment_table, aes(x = sex, y = count, fill = employed)) +
  geom_col(position = "dodge") + 
  labs(
    title = "Employment Status by Gender", 
    x = "Gender", 
    y = "Count", 
    fill = "Employment Status"
  ) + 
  theme_minimal()
```

```{r}
#| label: paper-figure-2
#Insurance and Employment visualization

insurance_by_employment <- wave5_clean |> 
  filter(!is.na(employed), !is.na(insurance_status)) |> 
  group_by(employed, insurance_status) |> 
  summarise(count = n(), .groups = "drop")

ggplot(insurance_by_employment, aes(
  y = fct_reorder(insurance_status, count), 
  x = count, 
  fill = employed)) +
  geom_col(position = "dodge") +
  labs(
    title = "Health Insurance Coverage by Employment Status", 
    x = "Count", 
    y = "Insurance Type", 
    fill = "Employment"
  ) + 
  theme_minimal()
```


```{r}
#| label: paper-table-2

# descriptive table of job type.

jobtype_table <- wave5_clean |> 
  filter(!is.na(job_type)) |> 
  count(job_type, name = "count") |> 
  mutate(percent = count/ sum(count))

jobtype_table |> 
  gt() |> 
  tab_header(title = "Distribution of Job Types Among Participants ") |> 
  cols_label(
    job_type = "Job Type", 
    count = "Count", 
    percent = "Percent"
  ) |> 
  fmt_percent(columns = percent, decimals = 1)
```



```{r}
#| label: paper-figure-3

#Health Insurance Status by Job Type
jobtype_insurance <- wave5_clean |> 
  filter(!is.na(job_type), !is.na(insurance_status)) |> 
  count(job_type, insurance_status) |> 
  group_by(job_type) |> 
  mutate(percent = n / sum(n)) |> 
  ungroup()

jobtype_insurance |> 
  gt() |> 
  tab_header(title = "Health Insurance Status by Job Type") |> 
  cols_label(
    job_type = "Job Type", 
    insurance_status = "Health Insurance", 
    n = "Count", 
    percent = "Percent"
  ) |> 
  fmt_percent(columns = percent, decimals = 1)
```


```{r}
#| label: paper-figure-2
#Figure: Health Insurance Coverage by Job Type 


jobtype_insurance |>  ggplot(aes(
  x = fct_reorder(job_type, percent, .fun = max), 
  y = percent, 
  fill = insurance_status
)) + 
  geom_col() +
  coord_flip() +
  labs(
    title = "Health Insurance Distribution by Job Type", 
    x = "Job Type", 
    y = "Percent Within Job Type", 
    fill = "Insurance Type"
  ) + 
  theme_minimal() + 
 scale_y_continuous(
   breaks = seq(0, 1, 0.2),
   labels = c("0%", "20%", "40%", "60%", "80%", "100%")
 )


```

```{r}
#| label: paper-figure-3


```

```{r}
#| label: paper-figure-4


```

```{r}
#| label: paper-figure-5


```

**code copied from data script draft that we can build off of for final paper, Delete at the end**

```{r}

# Data Exploratory visualization for Household Income Count

wave5_clean %>%
  filter(!is.na(household_income)) %>%
  group_by(household_income) %>%
  summarise(count = n()) %>%
  ggplot(aes(household_income, count)) +
  geom_col() +
  labs(title = "Count by Income Level", x = "Household Income Level", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


#Distribution of Job Types - Bar Chart
ggplot(wave5_clean, aes(x = job_type)) + 
  geom_bar() +
  coord_flip() + 
  labs(title = "Distribution of Job Types Among Participants", 
       x = "Job Type", 
       y = "Count")

#sex_employment plot
ggplot(sex_employment, aes(x = sex, y = n, fill = employed)) + 
  geom_col(position = "dodge", na.rm = TRUE) + 
  labs(title = "Employment Status by Sex", 
       x = "Sex",
       y = "Count",
       fill = "Employment")

# example of creating a subset of our data or a new data frame first then making tables or graphs
sex_employment <- wave5_clean |> 
  group_by (sex, employed) |>
  summarize(n = n(), .groups = "drop")



#hashtagged out for now to check rendering and citations

# table formatting
#gt() %>%
  #tab_header(title = "") %>%
  #cols_label(
    #household_income = "",
    #count = "",
    #percent = ""
  #) %>%
  #fmt_percent(columns = percent, decimals = 1) %>%
  #cols_align(align = "left", columns = household_income) %>%
  #cols_align(align = "right", columns = c(count, percent))
  
  


```

## Results

We used the following packages [@R; @here;@tidyverse;@rio;@janitor;@gt;@labelled;@webshot2] to assist our coding. 

## Discussion

### Findings about Gender and Employment
The descriptive table 1 and figures 1 highlight differences in employment status by gender and how it is linked to health insurance coverage. In this sample and analysis of Gender and Employment table, women have higher raw counts of employment because the female sample is larger.However, men shows the higher employment rate (87.3%) compared to women (79.6%). Women also have a higher count of "not employed" but proportionally still lower than men. Since there are small NA counts, it suggests that there will be minimal missing data and do not affect the patterns that much. This table of Gender and Employment status reflects labor-force disparities in U.S. It suggests slight gender differences in employment stability or participation rates.

The figure 1: Employment Status by Gender visualizes the same pattern that both genders are mostly employed. The bar chart visually shows larger absolute employment counts for women. However, proportionally, men are more likely to be employed when sample size is considered. It indicates gender differences in labor market participation in this sample. This analysis highlights the importance of distingushing between absolute counts and percentage. Employment is tightly linked to employer-sponsored insurance, so even small gender differences can influence coverage patterns. 

### Implications for Health Insurance
Higher male employment rates explain why men may have greater access to employer-based health insurance. Since women have relatively lower employment rate, it may increase reliance on non-employer insurance options such as Medicaid and partner insurance. This is because U.S. insurance is heavily employment-dependent, gendered labor-market patterns can translate directly into coverage disparities. These findings align with broader literature on women's higher likelihood of part-time work or employment interruptions which may limit access to employer benefits.

###Findings about Job Type
Job types vary significantly across participants and private for-profit employment is the most common one. Government jobs such as local, state and federal also represent meaningful shares of the sample. Self-employed and unpaid family workers represents smaller but important groups.

###Insurance Access Patterns
Private for-profit and government job types show the highest rates of employer-sponsored insurance while self-employed individuals have much lower access to employer insurance and rely more on individual market plans or remin uninsured. Unpaid or informal job types show very low employer-insurance coverage and higher Medicaid and uninsured rates. Therefore, job type is a strong structural determinant of insurance access even among people who are technically "employed". 

### Implications
Employer-provided insurance is unequally distributed across job sectors. Workers in traditional and stable employment sectors such as governments or for-profit benefit from higher coverage. Workers in self-employment or informal roles face systemic disadvantages in getting decent insurance. This support the argument that the structure of the labor market creates disparities in health insurance acess beyond employment status alone. 

## References



